Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-13 optim.qmd
minor text polishing
Minor polishing + fixed a few typos

* British English ("American or British usage is accepted, but
not a mixture of these")*

* No "we" - always third person

* Check if the affiliation information is correct

* word use:

  * Use jellyfish search optimiser, or JS optimiser, not jellyfish optimiser (my bad for leading the way)
  * "hyper-parameter" rather than "hyperparameter"

@@

The data structure proposed in @RJ-2021-105 is implemented for JS optimisers and  additional information such as index function used, data dimension, and jellyfish hyper-parameters can be included as additional columns. Below prints an example data collected from finding the sine-wave structure in 6D data (`d = 6`) using a spline index function(`splines2d`):

@@

-   **success rate**:  Success rate is defined as the proportion of simulations that achieve a final index value within 0.05 of the best index value found among all 50 simulations. In the dataset above, the optimal index value is `r sim_sine_6d_spline_best |> pull(index_val) |> max() |> round(3) -> aa; aa`. Out of the 50 simulations, `r sim_sine_6d_spline_best |> filter(index_val >= aa - 0.05) |> nrow() -> bb; bb` achieved an index value within [`r aa - 0.05`, `r aa`], resulting in a success rate of `r sim_sine_6d_spline_best |> mutate(a = index_val >= aa - 0.05) |> pull(a) -> cc; round(sum(cc)/length(cc), 3)`. @fig-projection shows the best projection found in each simulation, sorted by the index value. Most of the projections show a clear sine-wave structure, while the last six shows only a straight line with clusters at the end. When local optima exist in the projection pursuit problem, plotting the projections found in each simulation or by each jellyfish can help identify these local optima.

-  **Projection distance**: The projection distance between projection bases and the theoretical best basis (or the empirical best basis if the theoretical best is unknown) is calculated as the Frobenius norm, which measures how close a visited basis is to the best basis. Plotting this projection distance against the index value for each basis visited reveals how the index value changes as the basis approaches the optimal basis, helping to identify the squintability of the index function.  @fig-index-value-proj-dist shows the index value against the projection distance for the given example and its binned version. As the projection distance decreases, the index value initially increases with an increasing slope and then level off. The index value shows greater variation at medium projection distances compared to small and large distances.

@@

The JS optimiser provides a new swarm-based search strategy to optimise projection pursuit problems. Two examples are provided in this section to demonstrate its performance. The first example compares its performance with the search better optimiser and explores how it behaves with different combinations of hyper-parameter. The second example studies the effect of optimisation properties defined in @sec-theory, along with jellyfish hyper-parameters, on the outcome of the optimisation.

## Performance of JS optimiser in pipe-finding problems {#sec-app-1}

The performance of JS optimiser is investigated in two folds: 1) compare it with a commonly used optimiser: the search better optimiser [@RJ-2021-105; @laa_using_2020], and 2) examine its success rate under different hyper-parameters (number of jellyfishes and the maximum number of tries). The projection pursuit problem used is finding the pipe shape using the holes index, investigated by @laa_using_2020.

@@

Fifty (50) simulations are conducted with both JS and the search better optimiser, in four dimensions ($d = 6, 8, 10, 12$). The JS optimiser uses 100 jellyfishes and 100 maximum number of tries, while the search better optimiser allows a maximum of 1000 samples at each iteration before the algorithm terminates. @fig-proj presents the final projections found by the two optimisers, broken down by 10th quantile, faceted by the data dimension. In the 6-dimensional data scenario, the JS optimiser consistently identifies a clear pipe shape. The search better optimiser, although also find the pipe shape but with a wide rim, suggesting a further polish search may be required. With increasing dimensions, the JS optimiser may not always identify the pipe shape due to random sampling, it still presents a pipe shape in over 50% of cases. When compared to the search better optimiser, the JS optimiser reaches higher index values and clearer pipe shapes across all quantiles.

In the second experiment, fifty (50) simulations are conducted at each  hyper-parameter combination to analyse its effects on the success rate of the JS optimiser. The hyper-parameters tested include: 1) 20, 50, and 100 jellyfishes, and 2) 50 and 100 maximum number of tries. In each simulation,  success rate is calculated as defined in @sec-vis. @fig-proportion presents the proportion of success for each hyper-parameter combinations. As the number of jellyfish and maximum tries increase, the success rate also increases. For simpler problems (6 dimensions), small parameter values (20 jellyfishes and a maximum number of tries of 50) can already result in high success rate, while for higher-dimensional problems (8, 10, and 12 dimensions), a combination of 100 jellyfishes and 100 maximum number of tries is necessary for at least half of the jellyfishes to find the optimal projection.

## Factors affecting JS optimiser success rate: optimisation properties and jellyfish hyper-parameters {#sec-app-2}

The optimisation properties, including smoothness, squintability, and speed, offer numerical metrics to characterise the complexity of a projection pursuit optimisation problem. This example investigates how these metrics, alongside jellyfish hyper-parameters (the number of jellyfish and the maximum number of tries), affect the success of the JS optimiser. Simulations are conducted to obtain the success rate across various projection pursuit problems, characterised by shape-to-find, data dimension, and index function used, as well as different hyper-parameter combinations. Smoothness and squintability are calculated for each projection pursuit problem as outlined in @sec-theory. A generalised linear model is used to construct the relationship between the success rate, jellyfish hyper-parameters and optimisation properties.

In addition to the pipe-finding problem, a new shape, sine wave, is investigated in 6D and 8D spaces with six indices considered: `dcor2d_2`, `loess2d`, `MIC`, `TIC`, `spline`, and `stringy`. Combining with two jellyfish hyper-parameters, a total of `r nrow(sim_df)` cases is produced, comprising of `r nrow(filter(sim_df, index == "holes"))` pipe-finding cases  and `r nrow(filter(sim_df, index != "holes"))` sine-wave finding cases. For each case, the JS optimiser is run fifty (50) times and the success rate is calculated as in @sec-app-1.

*Note: should we print out the 52 cases investigated somewhere, say appendix, since it is not full grid simulation (otherwise for the sine-wave problem there will be 6 indexes x 6 configurations = 36 cases for 6-D data and 2 indexes [MIC & TIC] x 6 configurations = 12 cases for 8-D data)? We only do a selection of 28 of them to save time/ computational resources.*

Smoothness and squintability are computed for each case, following the procedures outlined in @sec-smoothness and @sec-squintability. To calculate smoothness, three hundred random bases are simulated. Index values and projection distance (to the optimal basis) are calculated for each random basis before fitting them into a Gaussian process model. For squintability, fifty random bases are sampled and interpolated to the optimal basis with a step size of 0.005. For these interpolated bases, index values and projection distances to the optimal basis are calculated and averaged with a bin width of 0.005. The squintability measure is then calculated from fitting a 4-parameter scaled logistic function of index value against projection distance, estimated by non-linear least square.

@@

@tbl-mod-data combines all the variables calculated above (the JS optimiser success rate, jellyfish hyper-parameters, and optimisation properties) into one table. Three pre-processing steps are applied to the data: 1) scaling the jellyfish hyper-parameters by a factor of 10 for interpretation, 2) creating a new binary variable `long_time` to indicate cases with an average run time over 30 seconds, and 3) encoding the success rate for the stringy case as 0, since none of the 50 simulations identified the sine-wave shape. A generalised linear model with a binomial family and a logit link function is used to fit the data and @tbl-mod-output presents the model outputs. The model suggests that the success rate of the JS optimiser is positively associated with the two jellyfish hyper-parameters, as well as with smoothness and squintability. However, being flagged with long runtime and an increase of data dimension will reduce the success rate. The variable `squintability` and `dimension` are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation success.

>>>>

-   British English ("American or British usage is accepted, but not a
    mixture of these")\*

-   No "we" - always third person

-   Check if the affiliation information is correct

-   word use:

    -   Use jellyfish search optimiser, or JS optimiser, not jellyfish
        optimiser (my bad for leading the way)
    -   "hyper-parameter" rather than "hyperparameter"

@@

The data structure proposed in @RJ-2021-105 is implemented for JS
optimisers and additional information such as index function used, data
dimension, and jellyfish hyper-parameters can be included as additional
columns. An example data collected from finding the sine-wave structure
in 6D data (`d = 6`) using a spline index function(`splines2d`) is
printed below:

@@

-   **Success rate**: The success rate is defined as the proportion of
    simulations that achieve a final index value within 0.05 of the best
    index value found among all 50 simulations. In the dataset above,
    the optimal index value is
    `r sim_sine_6d_spline_best |> pull(index_val) |> max() |> round(3) -> aa; aa`.
    Out of the 50 simulations,
    `r sim_sine_6d_spline_best |> filter(index_val >= aa - 0.05) |> nrow() -> bb; bb`
    achieved an index value within \[`r aa - 0.05`, `r aa`\], resulting
    in a success rate of
    `r sim_sine_6d_spline_best |> mutate(a = index_val >= aa - 0.05) |> pull(a) -> cc; round(sum(cc)/length(cc), 3)`.
    @fig-projection shows the best projection found in each simulation,
    sorted by the index value. Most of the projections show a clear
    sine-wave structure, while the last six show only a straight line
    with clusters at the end. When local optima exist in the projection
    pursuit problem, plotting the projections found in each simulation
    or by each jellyfish can help identify these local optima.

-   **Projection distance**: The projection distance between projection
    bases and the theoretical best basis (or the empirical best basis if
    the theoretical best is unknown) is calculated as the Frobenius
    norm, which measures how close a visited basis is to the best basis.
    Plotting this projection distance against the index value for each
    basis visited reveals how the index value changes as the basis
    approaches the optimal basis, helping to identify the squintability
    of the index function. @fig-index-value-proj-dist shows the index
    value against the projection distance for the given example and its
    binned version. As the projection distance decreases, the index
    value initially increases with an increasing slope and then levels
    off. The index value shows greater variation at medium projection
    distances compared to small and large distances.

@@

The JS optimiser provides a new swarm-based search strategy to optimise
projection pursuit problems. Two examples are provided in this section
to demonstrate its performance. The first example compares its
performance with the search-better optimiser and explores how it behaves
with different combinations of hyper-parameters. The second example
studies the effect of optimisation properties defined in @sec-theory,
along with jellyfish hyper-parameters, on the outcome of the
optimisation.

## Performance of the JS optimiser in pipe-finding problems {#sec-app-1}

The performance of the JS optimiser is investigated in two folds: 1)
compare it with a commonly used optimiser: the search-better optimiser
[@RJ-2021-105; @laa_using_2020], and 2) examine its success rate under
different hyper-parameters (number of jellyfishes and maximum number of
tries). The projection pursuit problem used is finding the pipe shape
using the holes index, investigated by @laa_using_2020.

@@

Fifty simulations are conducted with both JS and the search-better
optimiser, in four dimensions ($d = 6, 8, 10, 12$). The JS optimiser
uses 100 jellyfishes and 100 maximum number of tries, while the
search-better optimiser allows a maximum of 1000 samples at each
iteration before the algorithm terminates. @fig-proj presents the final
projections found by the two optimisers, broken down by 10th quantile,
faceted by the data dimension. In the 6-dimensional data scenario, the
JS optimiser consistently identifies a clear pipe shape. The
search-better optimiser also finds the pipe shape but with a wide rim,
suggesting a further polish search may be required. With increasing
dimensions, the JS optimiser may not always identify the pipe shape due
to random sampling, but it still presents a pipe shape in over 50% of
cases. When compared to the search-better optimiser, the JS optimiser
reaches higher index values and clearer pipe shapes across all
quantiles.

In the second experiment, fifty simulations are conducted at each
hyper-parameter combination to analyse its effects on the success rate
of the JS optimiser. The hyper-parameters tested include: 1) 20, 50, and
100 jellyfishes, and 2) 50 and 100 maximum number of tries. In each
simulation, the success rate is calculated as defined in @sec-vis.
@fig-proportion presents the proportion of success for each
hyper-parameter combinations. As the number of jellyfishes and maximum
tries increase, the success rate also increases. For simpler problems (6
dimensions), small parameter values (20 jellyfishes and a maximum number
of tries of 50) can already result in high success rate, while for
higher-dimensional problems (8, 10, and 12 dimensions), a combination of
100 jellyfishes and 100 maximum number of tries is necessary for at
least half of the jellyfishes to find the optimal projection.

## Factors affecting the JS optimiser success rate: optimisation properties and jellyfish hyper-parameters {#sec-app-2}

The optimisation properties, including smoothness, squintability, and
speed, offer numerical metrics to characterise the complexity of a
projection pursuit optimisation problem. This example investigates how
these metrics, along with the jellyfish search hyper-parameters (the
number of jellyfishes and the maximum number of tries), affect the
success of the JS optimiser. Simulations are conducted to obtain the
success rate across various projection pursuit problems, characterised
by shape-to-find, data dimension, and index function used, as well as
different hyper-parameter combinations. Smoothness and squintability are
calculated for each projection pursuit problem as outlined in
@sec-theory. A generalised linear model is used to construct the
relationship between the success rate, jellyfish hyper-parameters and
optimisation properties.

In addition to the pipe-finding problem, a new shape, sine wave, is
investigated in 6D and 8D spaces with six indices considered:
`dcor2d_2`, `loess2d`, `MIC`, `TIC`, `spline`, and `stringy`. Combining
with two jellyfish hyper-parameters, a total of `r nrow(sim_df)` cases
is produced, comprising of `r nrow(filter(sim_df, index == "holes"))`
pipe-finding cases and `r nrow(filter(sim_df, index != "holes"))`
sine-wave finding cases. For each case, the JS optimiser is run fifty
times and the success rate is calculated as in @sec-app-1.

*Note: should we print out the 52 cases investigated somewhere, say
appendix, since it is not full grid simulation (otherwise for the
sine-wave problem there will be 6 indexes x 6 configurations = 36 cases
for 6-D data and 2 indexes \[MIC & TIC\] x 6 configurations = 12 cases
for 8-D data)? We only do a selection of 28 of them to save time/
computational resources.*

Smoothness and squintability are computed for each case, following the
procedures outlined in @sec-smoothness and @sec-squintability. To
calculate smoothness, three hundred random bases are simulated. Index
values and projection distance (to the optimal basis) are calculated for
each random basis before fitting them into a Gaussian process model. For
squintability, fifty random bases are sampled and interpolated to the
optimal basis with a step size of 0.005. For these interpolated bases,
index values and projection distances to the optimal basis are
calculated and averaged with a bin width of 0.005. The squintability
measure is then calculated from fitting a 4-parameter scaled logistic
function of index value against projection distance, estimated by
non-linear least squares.

@@

@tbl-mod-data combines all the variables calculated above (the JS
optimiser success rate, jellyfish hyper-parameters, and optimisation
properties) into one table. Three preprocessing steps are applied to the
data: 1) scaling the jellyfish hyper-parameters by a factor of 10 for
interpretation, 2) creating a new binary variable `long_time` to
indicate cases with an average run time over 30 seconds, and 3) encoding
the success rate for the stringy case as 0, since none of the 50
simulations identified the sine-wave shape. A generalised linear model
with a binomial family and a logit link function is used to fit the data
and @tbl-mod-output presents the model outputs. The model suggests that
the success rate of the JS optimiser is positively associated with the
two jellyfish hyper-parameters, as well as with smoothness and
squintability. However, being flagged with long runtime and an increase
of data dimension will reduce the success rate. The variable
`squintability` and `dimension` are significant, suggesting their
importance relative to jellyfish hyper-parameters in the optimisation
success.
